---
title: Getting Started with wbstats A World Bank R Package
author: Jesse Piburn
date: '2018-01-08'
slug: getting-started-with-wbstats-a-world-bank-r-package
categories:
  - R
  - wbstats
tags:
  - R
  - wbstats
---

The [World Bank](http://www.worldbank.org/) is a tremendous source of global socio-economic data; spanning several decades and dozens of topics, it has the potential to shed light on numerous global issues. The `wbstats` R package provides access to this data.

This post is meant to serve as a reference for getting started with using `wbstats`. There are lots of things that aren't mentioned, particularly several of the `wb` arguments that can be changed. For a more detailed overview see the [Github READ ME](https://github.com/GIST-ORNL/wbstats) or [Introduction to the wbstats R-package Vignette](https://cran.r-project.org/web/packages/wbstats/vignettes/Using_the_wbstats_package.html) 

You can install:

The latest release version (`0.2`) from CRAN with
```{r, eval = FALSE}
install.packages("wbstats")
```

or

The latest development version from github with
```{r, eval = FALSE}
devtools::install_github("GIST-ORNL/wbstats")
```

`wbstats version 0.2` includes

* Uses version 2 of the World Bank API that provides access to more indicators and metadata than the previous API version
* Access to all annual, quarterly, and monthly data available in the API
* Support for searching and downloading data in multiple languages
* Access to the World Bank Data Catalog Metadata, providing among other information; update schedules and supported languages
* Ability to return `POSIXct` dates for easy integration into plotting and time-series analysis techniques
* Returns data in either long (default) or wide format for direct integration with packages like `ggplot2` and `dplyr`
* Support for Most Recent Value queries
* Support for `grep` style searching for data descriptions and names
* Ability to download data not only by country, but by aggregates as well, such as High Income or South Asia
* Ability to specify `countries_only` or `aggregates` when querying data

## Downloading data with `wb`

The `wb` function is how you request data from the API. The only thing you need to get started is which indicator(s) you want to download and for what time period. The `indicator` parameter takes a vector of indicatorIDs that correspond to the data you want to download. We'll mention how to find these IDs below
```{r}
library(wbstats)

# Population growth (annual %)
pop_data <- wb(indicator = "SP.POP.GROW", startdate = 2005, enddate = 2016)

head(pop_data)
```

Notice that the first "country" listed is `Arab World` which of course is not a country at all. The default value for the `country` parameter is a special value of `all` which as you might expect, returns data on the selected `indicator` for every available country or region. If you are interested in only some subset of countries or regions you can pass along the specific codes to the `country` parameter.

The country and region codes that can be passed to the `country` parameter correspond to the coded values from the `iso2c`, `iso3c`, `regionID`, `adminID`, and `incomeID` from the `countries` data frame in `wb_cachelist` or the return of `wbcache()` (more on that later). Any values from the above columns can mixed together and passed to the same call. You can also use the special value `country = "countries_only"` to return only values for actual countries
```{r}
# Population growth (annual %)
pop_data <- wb(country = "countries_only", indicator = "SP.POP.GROW", startdate = 2005, enddate = 2016)

head(pop_data)
```

To query indvidiual countries you can use their `iso2c` or `iso3c` codes.
```{r}
# Population growth (annual %)
pop_data <- wb(country = "US", indicator = "SP.POP.GROW", startdate = 2005, enddate = 2016)

head(pop_data)
```

Queries with multiple indicators return the data in a long data format by default
```{r}
pop_gdp_long <- wb(country = c("US", "NO"), indicator = c("SP.POP.GROW", "NY.GDP.MKTP.CD"),
                   startdate = 1971, enddate = 1971)

head(pop_gdp_long)
```

or a wide format if parameter `return_wide = TRUE`. Note that to necessitate a this transformation the `indicator` column is dropped.
```{r}
pop_gdp_wide <- wb(country = c("US", "NO"), indicator = c("SP.POP.GROW", "NY.GDP.MKTP.CD"),
                   startdate = 1971, enddate = 1971, return_wide = TRUE)

head(pop_gdp_wide)
```


## Search available data with `wbsearch`

`wbsearch` allows you to search for indicators that match a certain term. By default it searches for matching terms in both the name and description of the indicators.

```{r}

pop_vars <- wbsearch("Population Growth")

head(pop_vars)
```

From here you can select which indicators we want and pass their `indicatorID` into the `wb` function
```{r}
pop_vars <- wbsearch("Population Growth")

pop_var_ids <- pop_vars[1:3, "indicatorID"]


pop_data <- wb(country = "countries_only", indicator = pop_var_ids, startdate = 2005, enddate = 2016)

head(pop_data)
```

That is pretty much all you need to know to get started searching and downloading data. There are of course more things that can be done, but before we do that now is a good time to introduce out friend `wb_cachelist` 

## One list to rule them all `wb_cachelist`

For performance and ease of use, a cached version of useful information from the World Bank API is provided with the `wbstats` R-package. This data is called `wb_cachelist` and provides a snapshot of available countries, indicators, and other relevant information. `wb_cachelist` is by default the the source from which `wbsearch()` searches and the place `wb()` uses to do input sanity checks. The structure of `wb_cachelist` is as follows
```{r}
library(wbstats)

str(wb_cachelist, max.level = 1)
```

Inside the `wb_cachelist` is a data.frame for every major endpoint of the World Bank data API. Some of them such as `lending` and `income` are not as interesting as others, but for our purposes here we'll quickly highlight the `countries` and `indicators` data.frames.

### The `countries` data frame

This data.frame contains all of the geographic information for the locations that are available. This information is useful for finds country codes as well as joining back with any data you queried for groupinp and visualizing by columns such as `region` or `income` group.

```{r, eval=FALSE}
wb_geo <- wb_cachelist$countries

head(wb_geo, n = 5)
```

```{r, echo=FALSE, results='asis', warning=FALSE}
library(knitr)
library(kableExtra)
options(knitr.table.format = "html")

wb_geo <- wb_cachelist$countries

kable(head(wb_geo, n = 5)) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                font_size = 12) %>%
    scroll_box(extra_css = "overflow-x: auto;")
```


### The `indicators` data frame

This data.frame contains information such as the description and source of all indicators that are available for download. 
```{r, eval=FALSE}
wb_ind <- wb_cachelist$indicators

head(wb_ind, n = 5)
```

```{r, echo=FALSE, results='asis', warning=FALSE}
library(knitr)
library(kableExtra)
options(knitr.table.format = "html")

wb_ind <- wb_cachelist$indicators

kable(head(wb_ind, n = 5)) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                font_size = 11) %>%
    scroll_box(extra_css = "overflow-x: auto;")
```

### 

Earlier when we used the `wbsearch` function, it is actually searching through this `indicators` data.frame from the `wb_cachelist`. Now that we know a little more about what the function is doing we have a few more options available to us. For example, we can use the `fields` parameter to change which fields in the `indicators` data.frame to search through

```{r}
blmbrg_vars <- wbsearch("Bloomberg", fields = "sourceOrg")
head(blmbrg_vars)

```


### Accessing updated available data with `wbcache()`

For the most recent information on available data from the World Bank API `wbcache()` downloads an updated version of the information stored in `wb_cachelist`. `wb_cachelist` is simply a saved return of `wbcache(lang = "en")`. To use this updated information in  `wbsearch()` or `wb()`, set the `cache` parameter to the saved `list` returned from `wbcache()`. It is always a good idea to use this updated information to insure that you have access to the latest available information, such as newly added indicators or data sources.

```{r, eval = FALSE}
# default language is english
new_cache <- wbcache()

# if missing the cache parameter defaults to wb_cachelist
oil_vars <- wbsearch("Crude Oil", cache = new_cache)
```


## Plotting & Mapping with `wbstats`

Below are a few examples of using `ggplot2` and `leaflet` to create charts and maps using data from `wbstats`. But first, a useful option to know is the `POSIXct = TRUE` parameter.

### Using `POSIXct = TRUE`
The default format for the `date` column is not conducive to sorting or plotting when downloading  monthly or quarterly data. To address this, if `TRUE`, the `POSIXct` parameter adds the additional columns `date_ct` and `granularity`. `date_ct` converts the default date into a `Date` class. `granularity` denotes the time resolution that the date represents. This option requires the use of the package `lubridate (>= 1.5.0)`. If `POSIXct = TRUE` and `lubridate (>= 1.5.0)` is not available, a `warning` is produced and the option is ignored
```{r}
oil_data <- wb(indicator = "CRUDE_WTI", mrv = 10, freq = "M", POSIXct = TRUE)

head(oil_data)
```

### Plotting with `ggplot2`

The `POSIXct = TRUE` option makes plotting and sorting dates much easier. Here is an example of monthly oil prices using `ggplot2`
```{r vig-ex-posixct-1, fig.height = 4, fig.width = 7.5}
library(wbstats)
library(ggplot2)

oil_data <- wb(indicator = c("CRUDE_DUBAI", "CRUDE_BRENT", "CRUDE_WTI", "CRUDE_PETRO"),
               startdate = "2012M01", enddate = "2017M12", freq = "M", POSIXct = TRUE)

ggplot(oil_data) +
  geom_line(aes(x = date_ct, y = value, colour = indicator)) + 
  labs(title = "Crude Oil Price Comparisons",
       x = "Date",
       y = "US Dollars")
```

## Mapping `wbstats` data with `sf`

Currently, `wbstats` does not include any default geometries or spatial features. However, thanks to the fantastic [Simple Features](https://github.com/r-spatial/sf) R Package, we can easily add support 

```{r, warning=FALSE, message=FALSE}
library(wbstats)
library(dplyr)
library(sf)

# world country polygons 'medium' scale 
world_geo <- rnaturalearth::ne_countries(scale = 50, returnclass = "sf")


pop_data <- wb(country = "countries_only", 
               indicator = "SP.POP.GROW", 
               mrv = 1)
  
pop_geo <- left_join(world_geo, pop_data, by = c("iso_a2" = "iso2c"))

```

### Mapping with `ggplot2`

As of this writing, the version of `ggplot2` on CRAN (`2.2.1`) does not have support for `sf` objects. To take advantage of the latest functionality you'll need to download the development version of `ggplot2` from github.

Matt Strimas-Mackey has a really great overview of spatial data support in the tidyverse that goes into a lot more detail on using `sf` objects with `ggplot2`, `dplyr`, and the rest of the `tidyverse` [Here](http://strimas.com/r/tidy-sf/) so I won't go into anymore detail, but here is an example adapted from his post  

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(viridis)

ggplot(pop_geo) +
  geom_sf(aes(fill = value)) +
  scale_fill_viridis("value") +
  ggtitle("Population Growth (annual % growth)") +
  theme_bw()

```


### Example using `leaflet`

`leaflet` is a great package for online interactive maps in R. Here is the same map as above using `leaflet`

```{r leaflet-example, eval = TRUE, warning=FALSE}
library(leaflet)

pal <- colorNumeric("viridis", domain = pop_geo$value)

labels <- sprintf("<strong>%s</strong><br/>%s: %g%%",
                  pop_geo$name_long, pop_geo$indicator, round(pop_geo$value, 2)) %>% 
  lapply(htmltools::HTML)

l <- leaflet(pop_geo, height = 400, width = "100%") %>%
      setView(20,25, zoom = 1) %>%
      addTiles() %>%
      addPolygons(
        fillColor = ~pal(value),
        weight = 1,
        opacity = 1,
        color = "grey",
        fillOpacity = 0.7,
        highlight = highlightOptions(
          weight = 3,
          color = "#666",
          dashArray = "",
          fillOpacity = 0.7,
          bringToFront = TRUE),
        label = labels,
        labelOptions = labelOptions(
          style = list("font-weight" = "normal", padding = "3px 6px"),
          textsize = "15px",
          direction = "auto")) %>%
      addLegend(pal = pal, values = ~value, opacity = 0.9, 
                title = NULL,
                position = "bottomright",
                labFormat = labelFormat(suffix = "%"))

l
```

### Getting Started Indicators

Here are a few indicators that can help get you started with using the `wbstats` package

```{r, echo=FALSE}

common_ids <- c("SI.POV.GINI", "NY.GDP.MKTP.KD", "NY.GDP.MKTP.KD.ZG",
                "SP.POP.TOTL", "SP.POP.GROW", "SP.DYN.CBRT.IN", "SP.DYN.IMRT.IN",
                "EN.ATM.CO2E.PC", "EN.ATM.CO2E.KT", "SL.UEM.TOTL.ZS")

library(knitr)
library(kableExtra)
options(knitr.table.format = "html")

kable(wbstats::wb_cachelist$indicators[wbstats::wb_cachelist$indicators$indicatorID %in% common_ids, c("indicatorID", "indicator") ],
      row.names = FALSE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                font_size = 12) %>%
    scroll_box(extra_css = "overflow-x: auto;")

```

### Features Coming Soon
* Full metadata search, including country, country-series, and footnotes
* Better support for mapping with `sf`
* Addition of the [World Bank Projects API](https://data.worldbank.org/data-catalog/projects-portfolio)
* Suggest a feature on [Github](https://github.com/GIST-ORNL/wbstats/issues)
